Simulating High-Dimensional Multivariate Data

### 2020-10-31 040-monte-carlo-evaluation.Rmd
revise MC studies rev3; example rev1

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.

@@

Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
For all selected bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
Yet generally the accuracy seems adequate for most applications, at least in these limited simulation settings.

Of course, one could choose marginals that lead to poor matching by construction.
In partice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
See [Discusion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.

@@

The bivariate simulations provide a low-dimensional perspectitive.
But, of course, this is not `bigsimr`'s primary purpose.
We'd like the random vector methodolgy to scale to high-dimensional problems.

DESCRIBE WHAT WE DID. Do we need to redo this?

>>>>

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
We use values that are motivated by our RNA-seq data, namely the mean probability and sizes estimated from the data ( see [Example applications for our motivating data](examples) for estimation details): Bivariate Negative Binomial ( $prob_1 = prob_2 = `r prob`, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
Figure \@ref(fig:ch040-biNegBinPlot) `bigsimr` recovers the Pearson specified correlations for bivariate negative binomial.

@@

Taken together, the studies show our methology are accuracy across the entire range of possible correlation values for the rank-based dependency measures in continuous marginals.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
Table XX contains the average absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, especially near the correlation bounds.
Since the bivariate normal exactly matches all dependency measures in our scheme, we also report relative MSE of the non-normal cases to the normal simulations.

```{r ch040-biMSEtable}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom') )
allDat$type <- factor(allDat$type, levels = c( 'kendall', 'spearman', 'pearson' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( mae = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse
knitr::kable(
  tabMSE, booktabs = TRUE,
  caption = 'Average abolute error in matching the target dependency and  '
)
```

Generally, the accuracy seems adequate for most applications, at least in these limited simulation settings.
Of course, one could choose marginals that lead to poor matching by construction.
In partice, we recommend users to always evaluate the accuracy for their application, using methods similar to those presented above.
See [Discusion](discussion) for future directions for fast Pearson matching and discrete-specific modifications.

@@

With some assurance of the method's accuracy from the low-dimensional perspective in the bivariate setting, we now turn to assessing whether the `bigsimr` can scale to large dimesional problems. 
Now we seek evidence to determine whether the method can scale to higher dimensions in a practical time.
Using our motivating RNA-seq data, described in [Background](background), we filtered the orignal 20,501 genes to the high-expressing genes at increasing percentiles, $1, 5, 10, 15, 20, 25\%$, to obtain $d=\{206, 1026, 2051, 3076, 4101, 5127\}$ marginals and $\binom{d}{2}$ pairwise correlations at each setting.
For example, for $d=5127$ there are $13,140,501$ correlation coefficients.
We estimated the marginal negative binomial parameters and the correlation coefficients from the RNA-seq data to seed our simulations.
( See [Example applications for our motivating data](examples) for a detailed descrition of estimation).
Figure \@ref(fig:ch040-gpuVScpuFig) displays computation times using various high-performance settings to produce $B=10,000$ random vectors.
The Pearson simulations are much faster since the correlation conversion steps are avoided (pre-processing step; see [Algorithms](algorithms)), but as we known from above the accuracy will suffer slightly, especially near the negative boundary of the possible correlations.
Matching Spearman's correlation at larger $d$ gets costly if one wants to produce $B=10,000$ random vectors at many different simulation settings.
We recommend computing this type of simulations on a computing cluster (possibly using `rslurm`; see the [Advance Use](advanced-use)).

### 2020-10-31 index.Rmd
revise MC studies rev3; example rev1

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - name: Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
#output: distill::distill_article
abstract: |
  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
---

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - name: Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler
bibliography:
  - bigsimr.bib
  - packages.bib
site: bookdown::bookdown_site
#output: distill::distill_article
abstract: |
  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation matrices, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
---
